If the Null Fits, You Must Omit: Measurement Artifacts Undermine Group Polarization Replications

### 2025-10-29 GroupPolarizationStatmod/paper/ms.tex
finished corpus table
tables and mostly put in new symbology

@@

@@

We formalize this experimental design in the $N\to\infty$ case where the
distribution of ordinal measurements is created by integrating continuous latent
opinion distributions.  Latent opinions in the model are normally distributed with
mean $\mu_t$ and variance $\sigma_t$ for time $t$. We are concerned with only two
times, \emph{pre-} and \emph{post-} deliberation ($t=0$ and $t=T$, respectively).
For short we can write $\mu_\mathrm{pre}$ and $\sigma_\mathrm{pre}$ for
pre-deliberation mean and variance (and similar for \emph{post}-deliberation).  
The latent opinion of participant $i$ at time $t$ is drawn from a normal 
distribution in the model, 
\begin{equation} 
  o_{i,t} \sim p(o; \mu_t, \sigma_t) = \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution} 
\end{equation} 
\noindent 
Latent opinions are assumed to be normally distributed since they are unbounded 
aggregations of neurobiological activity.

The measurement process with $N\to\infty$ participants is simulated by 
integrating the latent opinion distribution between ordinal bin thresholds, 
of which there are $B+1$, where $B$ is the number of opinion bins, with sequential
integer values $b$. In the Moscovici and Zavalloni (1969) example, $b \in
\{-3,-2,\ldots,2,3\}$ and $B = 7$.
Bin $b$ is defined by its lower and upper continuous-valued thresholds, 
$\theta_{b-1}$ and
$\theta_{b}$, respectively. Thresholds are given by the following equations
(and visualized as dotted lines in Figure~\ref{fig:distros}).
\begin{equation}
\theta_b = \begin{cases}
  -\infty         & \text{ if } b=0 \\
  \infty          & \text{ if } b=B \\
  b + \frac{1}{2} & \text{ otherwise.}
\end{cases}
\end{equation}
\noindent
With thresholds defined, the frequency of ordinal measurment $b$ at time $t$ 
is (Figure~\ref{fig:distros})
\begin{equation}
  p(b;~\mu_t, \sigma_t, B) = \int_{\theta_{b-1}}^{\theta_b} p(o; \mu_t, \sigma_t)
  do.
  \label{eq:ordinal-frequency}
\end{equation}

We can now define spurious group polarization and give examples of when it occurs.
First, the observed group polarization opinion shift is
\begin{equation}
  g = \meanobs_\mathrm{post} - \meanobs_\mathrm{pre}~,
  \label{eq:group-polarization}
\end{equation}
\noindent
where the expected value of observed opinions at time $t$ is
\begin{equation}
  \meanobst = \frac{1}{B} \sum_{b=1}^B b \cdot p(b; \mu_t, \sigma_t, B).
  \label{eq:meanobs}
\end{equation}
\noindent

\emph{Spurious} group polarization is when mean observed opinions seem to increase in 
extremity when the latent mean is held constant. Call the collection of
latent null model distribution parameters $\lambda = (\mu, \sigma_{pre}, \sigma_{post})$.
The group polarization opinion shift can then be written as a function of $\lambda$,
$g(\lambda)$. Formally, then, spurious group polarization is plausible if 
there exists at least one $\lambda$ for which $g(\lambda) = g_{pub} > 0$, where
$g_{pub}$ is the published group polarization shift value.
This could be thought of as sampling the observed ordinal distributions 
$N \to \infty$, collecting ordinal measurements of samples from 
each latent distribution corresponding to each phase, pre- and post-deliberation.

Next, we explain how to find a set of latent null parameters, $\lambda$, that
generate spurious group polarization when naively measured on an ordinal scale.

@@

To identify plausibly false detections of group polarization, we searched numerically
for latent parameters $\lambda = (\mu, \sigmapre, and \sigmapost)$ that generate
spurious group polarization, as explained immediately above. We developed an *ad hoc*
hillclimbing-style algorithm that finds $\lambda$ that finds $\lambda$ by minimizing
the difference between published group polarization shifts $g_pub$ and calculated shifts,
$g_\lambda$.
calculated with Equation~\ref{eq:meanobs}.

The algorithm to find $\lambda$ holds $\mu$ constant while searching for $\sigmapre$
and $\sigmapost$ that generate observed $\meanobspre$ and $\meanobspost$, within a
tolerance of $\pm10^{-3}$ (Algorithm~\ref{alg:hillclimbing}). 
The algorithm requires a researcher to specify an initial
guess for $\lambda$ and auxiliary parameter settings, which generally vary between
experimental conditions.  Furthermore, the algorithm does not always produce
reasonable distributions with a given $\lambda$, and sometimes it does not produce
spurious group polarization.  Therefore, it is necessary to sanity check the latent
and ordinal opinion distributions for the $\lambda$ found by the algorithm. If all is
well, the experimental condition can be marked as plausibly spurious.  To tame the
complexity involved in data entry and model fitting for each of these sixty
experimental conditions, we developed a web application to input data from control
the algorithm and inspect its results (Figure~\ref{fig:webapp}). Web application code is
available online in our GitHub
repository\footnote{\url{https://github.com/mt-digital/gp-statmod/blob/main/GroupPolarizationStatmod/app.R}}.

@@

Note that so far, we have only identified cases where spurious group polarization plausibly
occurs in the asymptotic case where $N\to\infty$ ($N$ being the number of participants).
Experimental conditions yielding plausibly spurious group polarization warrant 
some suspicion. If there exists set of latent simple consensus 
parameters $\lambda$ that create binned (Equation~\ref{eq:ordinal-frequency})
opinion distributions that naively seem to be group polarized, then the means
of at least some samples from these distibutions with empirical numbers of 
participants would also appear to be shifted consistent with group
polarization. This process itself does not tell us the false detection rate.
We use the identified simple consensus parameters to define pre- and post-deliberation
latent opinion distributions in simulated trials of group polarization 
experiments.

@@

To calculate the false detection rate we begin by simulating 
pre- and post-deliberation 
opinions by drawing $N_e$ samples from normal latent opinion
distributions that generated spurious group polarization, i.e., with
parameters labelled $\lambda_e$ found in the hillclimbing step. Binning these
random values simulates observations. $N_e$ is the number
of participants in experimental condition $e$. Each trial $i$ of $N_e$
simulated pre- and post-deliberation observations is then inspected to see
if spurious group polarization occurred in the simulation. Spurious group
polarization is determined by first inferring the latent parameters of each
simulated experimental trial by fitting an ordered probit model~\cite{Liddell2018}. 
The inferred parameters are then used to calculate the effect size for that
trial, Cohen's $d$, defined as~\cite[p. 331]{Liddell2018}
\begin{equation}
  d = \frac{\mupost - \mupre}{\sqrt{\frac{1}{2}(\sigmapost^2 + \sigmapre^2)}}.
  \label{eq:cohens}
\end{equation}
\noindent
The family-wise error rate is, $\alpha_{e} = \Pr(D|\neg E)$, is the 
fraction of trials, 
with $d_{ei} > d^*$ for the $i^\text{th}$ trial for experimental
condition $e$, written
\begin{equation}
  \alpha_e = \frac{1}{N_T} \sum_i^{N_T} \Theta(d_{ei}; d^*),
  \label{eq:study_aggregate}
\end{equation}
\noindent
where
\[
  \Theta(d_{ei}; d^*) = 
    \begin{cases}
      1 & \text{ if } d_{ei} > d^* \\
      0 & \text{ otherwise.}
    \end{cases}
\]
and $N_T = 1000$ is the number of trials. For simplicity we assume that the sign of
$d_{ei}$ is positive, meaning a shift to greater extremisim. It is possible that
$d_{ei}$ could be negative, indicating a shift to lesser extremity. Indeed we
observe many simulations with a shift to lesser extremity. For now we do not
include these as indicating a ``significant'' effect, which we return to in the
Analysis below.

@@

% We used our formal distribution model of group polarization experiments
% to identify plausible latent simple consensus distributions that generate spurious
% group polarization in 
Of the remaining 57, we identified simple consensus $\lambda$ that generate spurious
group polarization in 54 experimental conditions. In the other three, we did
identify $\lambda$ that generate spurious group polarization, but the initial
simulated distributions were highly polarized. 
We rejected $\lambda$ generated for
two spurious group polarization from Schkade, et al., (2010) because
the histogram plots of the observed opinion distributions are not
polarized~\cite[Figure 1, p. 234]{Schkade2010}. The other condition where we
found but rejected spurious group polarization $\lambda$ was in a question 
about a ``bad teacher'' in Myers (1975)~\cite{Myers1975a}, where lower values
signify increasingly worse impressions of the teacher and higher values signify
increasingly good impressions. It is unlikely that a
``bad teacher'' would be rated as good in this case, so we excluded this 
from our analysis as well. This exclusion may be overly generous since Myers (1975)
did not provide details beyond mean values and opinion shifts, and the article was
published before open data practices. Our corpus had a total of 60 experimental conditions, but we excluded 3 out of 60 
that did not claim to be positive findings of group polarization.

>>>>

\usepackage{floatrow}
\DeclareFloatFont{small}{\small}% "scriptsize" is defined by floatrow, "tiny" not
\floatsetup[table]{font=small}

\usepackage[table]{xcolor}
\usepackage{multirow}
\usepackage{makecell}
\usepackage{array}

@@

\begin{table}[h]
  \caption{\textbf{Latent psychological variables and distributions.}}
  \begin{tabular}{cll} 
  \toprule
    \textbf{Variable} & \textbf{Description} & \textbf{Values} \\
  \midrule  
    $\omega_{i,t}$ & Latent opinion of individual $i$ at time $t$ & $\mathbb{R}$ \\
    $\mu_t$ & Mean latent opinion at time $t$ (constant in null-polarization) & $\mathbb{R}$ \\
    $\sigma_t$ & Variance of observed opnions at time $t$ & $\mathbb{R}$ \\
  \bottomrule
  \end{tabular} 
\end{table}

\begin{table}[h]
  \caption{\textbf{Experimental design and numerical model parameters.}}
  \begin{tabular}{cll} 
  \toprule
    \textbf{Variable} & \textbf{Description} & \textbf{Values} \\
  \midrule  
    $B$ & Number of bins & $5,6,7,\ldots$ \\
    $\theta_b$ & The $B+1$ threshold values that separate bins & 
      e.g., $-\infty, -,\ldots$ \\
    $b$ & Bin index & $0,1,2,\ldots$ \\
  \bottomrule
  \end{tabular} 
\end{table}

\begin{table}[h]
  \caption{\textbf{Simulated observed variables.}}
  \begin{tabular}{cll} 
  \toprule
    \textbf{Variable} & \textbf{Description} & \textbf{Values} \\
  \midrule  
    $\meanobst$ & Mean of all reported opinions (Eq. \ref{eq:meanobs}) & $\mathbb{R}$ \\
    $g$ & Observed group opinion polarization (Eq. \ref{eq:group-polarization})
        & $\mathbb{R}$ \\
    $d$ & Cohen's $d$ used as significance threshold 
        & $\mathbb{R}$ \\
    $\alpha_e,\alpha_s,\alpha_{\text{all}}$ & Family-wise/Type I error rate
    for an experiment, study, or across all experiments
             & $[0, 1]$ \\
    FDR & False discovery rate
        & $[0, 1]$ \\
  \bottomrule
  \end{tabular} 
\end{table}

@@

We formalize this experimental design in the $N\to\infty$ case where the
distribution of ordinal measurements is created by integrating continuous latent
opinion distributions.  Latent opinions in the model are normally distributed with
mean $\mu_t$ and variance $\sigma_t$ for time $t$. We are concerned with only two
times, \emph{pre-} and \emph{post-} deliberation ($t=0$ and $t=T$, respectively).
For short we can write $\mu_\mathrm{pre}$ and $\sigma_\mathrm{pre}$ for
pre-deliberation mean and variance (and similar for \emph{post}-deliberation).  
The latent opinion of participant $i$ at time $t$ is drawn from a normal 
distribution in the model, 
\begin{equation} 
  \omega_{i,t} \sim p(\omega; \mu_t, \sigma_t) = \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution} 
\end{equation} 
\noindent 
Latent opinions are assumed to be normally distributed since they are unbounded 
aggregations of neurobiological activity.

The measurement process with $N\to\infty$ participants is simulated by 
integrating the latent opinion distribution between ordinal bin thresholds, 
of which there are $B+1$, where $B$ is the number of opinion bins, with sequential
integer values $b \in \{0, B\}$. In the Moscovici and Zavalloni (1969)
example, the thresholds are \in
\{-\infty,-2.5,-1.5,\ldots,2.5,\infty\}$, with $B = 7$.
Bin $b$ is defined by its lower and upper continuous-valued thresholds, 
$\theta_{b-1}$ and
$\theta_{b}$, respectively. Thresholds are given by the following equations
(visualized as dotted lines in Figure~\ref{fig:distros}).
\begin{equation}
\theta_b = \begin{cases}
  -\infty         & \text{ if } b=0 \\
  \infty          & \text{ if } b=B \\
  \theta_{\text{Base}} + \frac{b}{2} & \text{ otherwise.}
\end{cases}
\end{equation}
\noindent
In the Moscovici and Zavalloni example, the base bin value is
$\theta_{\text{Base}} = -3$.
With thresholds defined, the frequency of ordinal measurment $b$ at time $t$ 
is (Figure~\ref{fig:distros})
\begin{equation}
  p(b;~\mu_t, \sigma_t, B) = \int_{\theta_{b-1}}^{\theta_b} p(\omega; \mu_t, \sigma_t)
  d\omega.
  \label{eq:ordinal-frequency}
\end{equation}

We can now define spurious group polarization and give examples of when it
occurs. First, the expected value of observed opinions at time $t$ is
\begin{equation}
  \meanobst = \frac{1}{B} \sum_{b=1}^B b \cdot p(b; \mu_t, \sigma_t, B).
  \label{eq:meanobs}
\end{equation}
\noindent
The observed group polarization opinion shift is then
\begin{equation}
  g = \meanobs_\mathrm{post} - \meanobs_\mathrm{pre}.
  \label{eq:group-polarization}
\end{equation}
\noindent

Measurement artifacts can make null-polarization spuriously appear like group
polarization. To see how, we will build up a formal model of the
process. Consider a collection of
null-polarization latent opinion distribution parameters $\nu = (\mu, \sigma_{pre}, \sigma_{post})$.
The group polarization opinion shift can then be written as a function of
$\nu$,
$g(\nu)$.

If we can find ..... $\nu$ for which $g(\nu) = g_{pub} > 0$, where
$g_{pub}$ is the published group polarization shift value.
This could be thought of as sampling the observed ordinal distributions 
$N \to \infty$, collecting ordinal measurements of samples from 
each latent distribution corresponding to each phase, pre- and post-deliberation.

Next, we explain how to find a set of latent null parameters, $\nu$, that
generate spurious group polarization when naively measured on an ordinal scale.

@@

To identify plausibly false detections of group polarization, we searched numerically
for latent parameters $\nu = (\mu, \sigmapre, and \sigmapost)$ that generate
spurious group polarization, as explained immediately above. We developed an *ad hoc*
hillclimbing-style algorithm that finds $\nu$ that finds $\nu$ by minimizing
the difference between published group polarization shifts $g_pub$ and calculated shifts,
$g_\nu$.
calculated with Equation~\ref{eq:meanobs}.

The algorithm to find $\nu$ holds $\mu$ constant while searching for $\sigmapre$
and $\sigmapost$ that generate observed $\meanobspre$ and $\meanobspost$, within a
tolerance of $\pm10^{-3}$ (Algorithm~\ref{alg:hillclimbing}). 
The algorithm requires a researcher to specify an initial
guess for $\nu$ and auxiliary parameter settings, which generally vary between
experimental conditions.  Furthermore, the algorithm does not always produce
reasonable distributions with a given $\nu$, and sometimes it does not produce
spurious group polarization.  Therefore, it is necessary to sanity check the latent
and ordinal opinion distributions for the $\nu$ found by the algorithm. If all is
well, the experimental condition can be marked as plausibly spurious.  To tame the
complexity involved in data entry and model fitting for each of these sixty
experimental conditions, we developed a web application to input data from control
the algorithm and inspect its results (Figure~\ref{fig:webapp}). Web application code is
available online in our GitHub
repository\footnote{\url{https://github.com/mt-digital/gp-statmod/blob/main/GroupPolarizationStatmod/app.R}}.

@@

Note that so far, we have only identified cases where spurious group polarization plausibly
occurs in the asymptotic case where $N\to\infty$ ($N$ being the number of participants).
Experimental conditions yielding plausibly spurious group polarization warrant 
some suspicion. If there exists set of latent simple consensus 
parameters $\nu$ that create binned (Equation~\ref{eq:ordinal-frequency})
opinion distributions that naively seem to be group polarized, then the means
of at least some samples from these distibutions with empirical numbers of 
participants would also appear to be shifted consistent with group
polarization. This process itself does not tell us the false detection rate.
We use the identified simple consensus parameters to define pre- and post-deliberation
latent opinion distributions in simulated trials of group polarization 
experiments.

@@

To calculate the false detection rate we begin by simulating 
pre- and post-deliberation 
opinions by drawing $N_e$ samples from normal latent opinion
distributions that generated spurious group polarization, i.e., with
parameters labelled $\nu_e$ found in the hillclimbing step. Binning these
random values simulates observations. $N_e$ is the number
of participants in experimental condition $e$. Each trial $i$ of $N_e$
simulated pre- and post-deliberation observations is then inspected to see
if spurious group polarization occurred in the simulation. Spurious group
polarization is determined by first inferring the latent parameters of each
simulated experimental trial by fitting an ordered probit model~\cite{Liddell2018}. 
The inferred parameters are then used to calculate the effect size for that
trial, Cohen's $d$, defined as~\cite[p. 331]{Liddell2018}
\begin{equation}
  d = \frac{\mupost - \mupre}{\sqrt{\frac{1}{2}(\sigmapost^2 + \sigmapre^2)}}.
  \label{eq:cohens}
\end{equation}
\noindent
The family-wise error rate is, $\alpha_{e} = \Pr(D|\neg E)$, is the 
fraction of trials, 
with $d_{ei} > d^*$ for the $i^\text{th}$ trial for experimental
condition $e$, written
\begin{equation}
  \alpha_e = \frac{1}{N_T} \sum_i^{N_T} \Theta(d_{ei}; d^*),
  \label{eq:study_aggregate}
\end{equation}
\noindent
where
\[
  \Theta(d_{ei}; d^*) = 
    \begin{cases}
      1 & \text{ if } d_{ei} > d^* \\
      0 & \text{ otherwise.}
    \end{cases}
\]
and $N_T = 1000$ is the number of trials. For simplicity we assume that the sign of
$d_{ei}$ is positive, meaning a shift to greater extremisim. It is possible that
$d_{ei}$ could be negative, indicating a shift to lesser extremity. Indeed we
observe many simulations with a shift to lesser extremity. For now we do not
include these as indicating a ``significant'' effect, which we return to in the
Analysis below.

@@

% We used our formal distribution model of group polarization experiments
% to identify plausible latent simple consensus distributions that generate spurious
% group polarization in 
Of the remaining 57, we identified simple consensus $\nu$ that generate spurious
group polarization in 54 experimental conditions. In the other three, we did
identify $\nu$ that generate spurious group polarization, but the initial
simulated distributions were highly polarized. 
We rejected $\nu$ generated for
two spurious group polarization from Schkade, et al., (2010) because
the histogram plots of the observed opinion distributions are not
polarized~\cite[Figure 1, p. 234]{Schkade2010}. The other condition where we
found but rejected spurious group polarization $\nu$ was in a question 
about a ``bad teacher'' in Myers (1975)~\cite{Myers1975a}, where lower values
signify increasingly worse impressions of the teacher and higher values signify
increasingly good impressions. It is unlikely that a
``bad teacher'' would be rated as good in this case, so we excluded this 
from our analysis as well. This exclusion may be overly generous since Myers (1975)
did not provide details beyond mean values and opinion shifts, and the article was
published before open data practices. Our corpus had a total of 60 experimental conditions, but we excluded 3 out of 60 
that did not claim to be positive findings of group polarization.
